Analytical Uncertainty Quantification for Multilevel Mediation Analysis

### 2024-07-17 papers/Manuscript - Exact Asymptotics.tex
Delete old discussion of mediation effects identification

\newcommand{\zdens}{\frac{e^{-z^2/2}}{\sqrt{2\pi}}}

@@

One extension which is of particular interest is to dependent data via multilevel, or mixed-effects, models. Mixed-effects regression models involve the introduction of random, unobserved coefficients to an existing regression \hl{(awk)}. A common setting in which such a model arises is clustered data, where the random coefficients differ across clusters, but are constant within each cluster. See, e.g., \citet{Dem04} for an overview of mixed-effects methodology. When applied to mediation analysis, mixed-effects methods allow for modelling group-specific mediation effects, and the effect of this heterogeneity on the estimation of global effects.

In this paper, we present a general framework for multilevel mediation analysis based on the estimation of nested counterfactuals. In particular, our method can be applied to estimate mediation effects on whatever scale is of interest (e.g., risk difference, risk ration, odds ratio).

@@

\begin{outline}
    \1 Define counterfactuals (ctfs) and nested counterfactuals 
    \1 Define mediation effects
        \2 Continuous outcome
        \2 Binary outcome on risk difference, risk ratio and odds ratio scales
    \1 Identification of mediation effects
        \2 See \citet{Ima10I} for non-parametric identification of nested counterfactuals
    \1 Regression modelling
        \2 Restrict to binary outcome?
        \2 Fixed effects regression (brief)
        \2 Mixed effects regression
        \2 Prediction of random effects
\end{outline}

@@

Our approach to causal mediation analysis is based on the counterfactual framework of ????. Briefly, let $Y$ be an outcome of interest, $X$ be an exposure which is a causal driver of $Y$, and $M$ be a mediator, which influences $Y$ and is influenced by $X$. Figure \ref{fig:simp_med} shows a causal diagram representing this relationship. We call the top arrow the direct effect of $X$ on $Y$, and the bottom path through $M$ the indirect effect of $Y$. Taken together, these two pathways constitute the total effect of $X$ on $Y$.

@@

There are three main types of mediation effect: total, direct and indirect, although the scale on which these are measured can vary. For concreteness, we focus here only on discrepancies defined as differences; i.e., those of the form $Y_{x_1, M_{x_1'}} - Y_{x_2, M_{x_2'}}$. Extending our results to effects defined on different scales (e.g. ratios) is straightforward. We will return briefly to this in Section ????.

Proceeding now to the actual definitions, we define the total effect of $X$ on $Y$ to be $TE(x, x') = EY_x - EY_{x'}$. The direct effect is defined as $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect is $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$. See, e.g., ???? for motivation and discussion of these definitions.

@@

A fundamental problem of causal inference is that we can only ever observe one counterfactual outcome on a particular individual. In mediation analysis, this problem is even worse, since many of our definitions involve the nested counterfactual, $Y_{x, M_{x'}}$, which when $x \neq x'$ cannot be observed on any individual. Nevertheless, \citet{Ima10I} give conditions under which the population average of a nested counterfactual can be expressed in terms of conditional expectations, possibly conditional on one or more additional covariates, $W$. Specifically, their Theorem 1 states that, under a condition they call ``Sequential Ignorability'', we can write
%
\begin{equation}
    \bE (Y_{x, M_{x'}} | W=w) = \bE_M \left[ \bE_Y \left( Y | X=x, M, W=w \right) | X=x', W=w \right]. \label{eq:ID}
\end{equation}
%
In fact, their Theorem 1 is somewhat more general, giving an expression for the density of the nested counterfactual rather than its expected value.

@@

We extend this regression modelling with the introduction of random effects \citep[see, e.g.][]{Dem04}. Specifically, we include random effects for the intercept and $X$ in our model for $M$, and for the intercept, $X$ and $M$ in our model for $Y$. Let $U$ and $V$ be the random effects for our models of $Y$ and $M$ respectively. We can re-write Equation (\ref{eq:ID_bin}) as
%
\begin{align}
    \bE (Y_{x, M_{x'}} | W=w) = & \left[\bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) \right. \cdot \label{eq:ID_bin_ran}\\
    &  \left. \bE_V \bP(M=1 |V, X = x', W=w)\right] + \nonumber \\ 
    & \left[ \bE_U \bP \left( Y=1 |U, X=x, M=0, W=w \right) \right. \cdot \nonumber \\
    & \left. \bE_V \bP(M=0 |V, X = x', W=w) \right]. \nonumber 
\end{align}
%
Following the usual approach, we model the random effects $U$ and $V$ as normally distributed, allowing for correlation between effects from the same model, but assuming independence between models. Let $U \sim N(0, \Gamma_Y)$ and $V \sim N(0, \Gamma_M)$.

@@

Uncertainty quantification for our estimator, $\hat{\kappa}$, proceeds by the $\delta$-method \citep[see, e.g.,][]{vdV98}. We first get asymptotic covariance matrices for estimators of our two sets of regression parameters. Next, we stack these two covariances into a single $2\times 2$ block-diagonal matrix (\hl{assuming asymptotic independence between models, see} \citealp{Bau06}), then pre- and post-multiply by the Jacobian of $\kappa$. From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers. These objective functions differ from likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from the joint likelihood between the observed data and random effects.

We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package (\texttt{MultMedUQ}?). See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$. We will often also need the limiting covariance of $\kappa$ evaluated at two sets of values for $x$ and $x'$. In this case we write $\rho(x_1, x'_1; x_2, x'_2)$ for the asymptotic correlation between $\hat{\kappa}(x_1, x'_1)$ and $\hat{\kappa}(x_2, x'_2)$.

\subsection{Mediation Effects}

@@

\subsection{Old}

In this section, we define the total, direct and indirect mediation effects in terms of nested counterfactuals. We begin by applying the Mediation Formula of \citet{Pea12} to identify nested counterfactuals with simple functions of conditional expectations.

We begin by observing that mediation effects are often defined in terms of nested counterfactuals (regardless of the scale on which these effects are reported). \hl{There is some causal inference theory to be done here. For now, I'm just going to write what I expect to be true, then later I will go back and add the necessary assumptions.} As such, we begin by identifying a general nested counterfactual with estimable quantities. To this end, write $Y(x, m)$ for the counterfactual value of $Y$ when $X$ and $M$ are set to $x$ and $m$ respectively. Similarly, write $M(x)$ for the counterfactual value of $M$ when $X=x$. We write $Y(x, M(x'))$ for the value of $Y$ when $X$ is set to $x$ and $M$ is set to what it would have been if $X$ were $x'$. We refer to $Y(x, M(x'))$ as a nested counterfactual. Note that, when $x \neq x'$, the nested counterfactual is necessarily unobservable. Nevertheless, given regularity conditions, we can write
%
\begin{equation}
    \bE Y(x, M(x')) = \int \bE(Y | M=m, X=x) \bP(M = dm | X = x') \label{eq:id_nest}
\end{equation}
%
A similar expression holds conditional on pre-treatment covariates,
\begin{equation}
    \bE Y_c(x, M(x')) = \int \bE(Y | M=m, X=x, C=c) \bP(M = dm | X = x', C=c) \label{eq:id_nest_cov}
\end{equation}
%
Having identified expected nested counterfactuals with integrals of conditional quantities, we can use regression to estimate the RHS of Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov}. This identification with regression-based quantities is especially simple if $M$ is binary. Here, we get $\bP(M=m | \cdot) = \bE(M=m | \cdot)$, where the latter quantity is popularly modelled with logistic regression. For simplicity, we hereafter assume that the mediator, $M$, is binary. \hl{If not, we need to do something different. One option is to extract the conditional density of $(M | \cdot)$, then perform the integral. Another is to do regression with response $\bE(Y | M=m, X=x, C=c)$, viewed as a function of $m$. There is likely an equivalence between these two methods; I can come back to this later.}

The above development allows only fixed-effects in the regression models. If we want to also incorporate random effects, we simply view Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov} as having had these random effects marginalized out. That is, if we write $U$ and $V$ for the random effects in our models for $Y$ and $M$ respectively, and $G_V$, $G_V$ for the random effects' distributions, then we can re-write Equations \ref{eq:id_nest} and \ref{eq:id_nest_cov} as
%
\begin{align}
    \bE Y(x, M(x')) &= \int \bE(Y | U=u, M=m, X=x) \bP(M = dm |V=v, X = x') G_U(du) G_V(dv) \label{eq:ran_id_nest} \\
    \bE Y_c(x, M(x')) &= \int \bE(Y | U=u, M=m, X=x, C=c) \bP(M = dm | V=v, X = x', C=c) G_U(du) G_V(dv) \label{eq:ran_id_nest_cov}
\end{align}

\subsection{Ideas for CI Theory}
Applying the Counterfactual Unnesting Theorem of \citet{Cor21}, and provided that the conditions specified therein are satisfied, we have

Alternatively, see Theorem 1 of \citet{Ima10I}. This is specifically the result I want, not the corresponding identification results of \citet{Ima10II}, which focuses more narrowly on identification of mediation effects. Note that, in \citet{Ima10I}, the function $f$ is a density (with respect to some unspecified measure, possibly counting measure). In order to get the above formulas for expected values of $Y$, we do need to assume that $\bE Y$ is finite so we can apply Fubini's Theorem (the one for finite integrals).

@@

\begin{outline}
    \1 Estimate models using \texttt{lme4}
    \1 UQ for model parameters (mention \texttt{merDeriv} package)
    \1 UQ for nested counterfactuals via $\delta$-method
    \1 Simultaneous UQ for triples of mediation effects (i.e. total, direct and indirect)
        \2 Can also do all 9 simultaneously
    \1 UQ for predicted group-level effects?
\end{outline}

@@

\begin{outline}
    \1 Monte Carlo study
        \2 Proof of concept
        \2 Explore robustness. See \citet{Sam23} for inspiration.
    \1 Real data
        \2 Trust study dataset?
\end{outline}

>>>>

\newcommand{\zdens}{\frac{e^{-z^2/2}}{\sqrt{2\pi}}}
\newcommand{\indep}{\perp \!\!\! \perp}

@@

One extension which is of particular interest is to dependent data via multilevel, or mixed-effects, models. Mixed-effects regression models involve the introduction of random, unobserved coefficients to an existing regression. A common setting in which such a model arises is clustered data, where the random coefficients differ across clusters, but are constant within each cluster. See, e.g., \citet{Dem04} for an overview of mixed-effects methodology. When applied to mediation analysis, mixed-effects methods allow for modelling group-specific mediation effects, and the effect of this heterogeneity on the estimation of global effects.

In this paper, we present a general framework for multilevel mediation analysis based on the estimation of nested counterfactuals. In particular, our method can be applied to estimate mediation effects on whatever scale is of interest (e.g., risk difference, risk ration, odds ratio). We focus our presentation on a binary outcome and binary mediator, although the extension to one or both being continuous is straightforward.

@@

\begin{outline}
    \1 \st{Define counterfactuals (ctfs) and nested counterfactuals }
    \1 \st{Define mediation effects}
        \2 \st{Continuous outcome}
        \2 \st{Binary outcome on risk difference, risk ratio and odds ratio scales}
    \1 \st{Identification of mediation effects}
        \2 \st{See} \citet{Ima10I} \st{for non-parametric identification of nested counterfactuals}
    \1 \st{Regression modelling}
        \2 \st{Restrict to binary outcome?}
        \2 \st{Fixed effects regression (brief)}
        \2 \st{Mixed effects regression}
        \2 \st{Prediction of random effects}
\end{outline}

@@

Our approach to causal mediation analysis is based on the well-known counterfactual framework \citep[see, e.g.,][]{Van09,Pea12}. Briefly, let $Y$ be an outcome of interest, $X$ be an exposure which is a causal driver of $Y$, and $M$ be a mediator, which influences $Y$ and is influenced by $X$. Figure \ref{fig:simp_med} shows a minimal causal diagram representing this relationship. We call the top arrow the direct effect of $X$ on $Y$, and the bottom path through $M$ the indirect effect of $Y$. Taken together, these two pathways constitute the total effect of $X$ on $Y$.

@@

There are three main types of mediation effect: total, direct and indirect, although the scale on which these are measured can vary. For concreteness, we focus here only on effects defined as differences; i.e., those of the form $Y_{x_1, M_{x_1'}} - Y_{x_2, M_{x_2'}}$. Extending our results to effects defined on different scales (e.g. ratios) is straightforward; see Section \ref{sec:med_effs}.

Proceeding now to the definitions, we set the total effect of $X$ on $Y$ to be $TE(x, x') = EY_x - EY_{x'}$. The direct effect is $DE(x, x') = Y_{x, M_{x'}} - Y_{x', M_{x'}}$, and the indirect effect is $IE(x, x') = Y_{x, M_{x}} - Y_{x, M_{x'}}$. See, e.g., \citet{Pea12} for motivation and discussion of these definitions.

@@

A fundamental problem of causal inference is that we can only ever observe one counterfactual outcome on a particular individual. In mediation analysis, this problem is even worse, since many of our definitions involve the nested counterfactual, $Y_{x, M_{x'}}$, which when $x \neq x'$ cannot be observed on any individual. Nevertheless, \citet{Ima10I} give conditions under which the population average of a nested counterfactual can be expressed in terms of conditional expectations, possibly conditional on one or more additional covariates, $W$. Specifically, their Theorem 1 states that, under a condition they call ``Sequential Ignorability'', we can write
%
\begin{equation}
    \bE (Y_{x, M_{x'}} | W=w) = \bE_M \left[ \bE_Y \left( Y | X=x, M, W=w \right) | X=x', W=w \right]. \label{eq:ID}
\end{equation}
%
In fact, their Theorem 1 is somewhat more general, giving an expression for the density of the nested counterfactual rather than its expected value. We henceforth assume that Sequential Ignorability is satisfied by the variables in our model.

@@

We extend this regression modelling with the introduction of random effects \citep[see, e.g.][]{Dem04}. Specifically, we include random effects for the intercept and $X$ in our model for $M$, and for the intercept, $X$ and $M$ in our model for $Y$. Let $U$ and $V$ be the random effects for our models of $Y$ and $M$ respectively. We can re-write Equation (\ref{eq:ID_bin}) as
%
\begin{align}
    \bE (Y_{x, M_{x'}} | W=w) = & \left[\bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) \right. \cdot \label{eq:ID_bin_ran}\\
    &  \left. \bE_V \bP(M=1 |V, X = x', W=w)\right] + \nonumber \\ 
    & \left[ \bE_U \bP \left( Y=1 |U, X=x, M=0, W=w \right) \right. \cdot \nonumber \\
    & \left. \bE_V \bP(M=0 |V, X = x', W=w) \right]. \nonumber 
\end{align}
%
We model the random effects $U$ and $V$ as normally distributed, allowing for correlation between effects from the same model, but assuming independence between models (i.e. $U \indep V$). Let $U \sim N(0, \Gamma_Y)$ and $V \sim N(0, \Gamma_M)$.

@@

Uncertainty quantification for our estimator, $\hat{\kappa}$, proceeds by the $\delta$-method \citep[see, e.g.,][]{vdV98}. We first get asymptotic covariance matrices for estimators of our two sets of regression parameters. Next, we stack these two covariances into a single $2\times 2$ block-diagonal matrix (\hl{assuming asymptotic independence between models, see} \citealp{Bau06}), then pre- and post-multiply by the Jacobian of $\kappa$. From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers. These objective functions differ from log-likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from its joint likelihood with the random effects.

We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package\footnote{Name idea: \texttt{MultMedUQ}?}. See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$. We will often also need the limiting covariance of $\kappa$ evaluated at two sets of values for $x$ and $x'$. In this case we write $\rho(x_1, x'_1; x_2, x'_2)$ for the asymptotic correlation between $\hat{\kappa}(x_1, x'_1)$ and $\hat{\kappa}(x_2, x'_2)$.

\subsection{Mediation Effects}
\label{sec:med_effs}

@@

@@

\begin{outline}
    \1 \st{Estimate models using \texttt{lme4}}
    \1 \st{UQ for model parameters (mention \texttt{merDeriv} package)}
    \1 \st{UQ for nested counterfactuals via $\delta$-method}
    \1 \st{Simultaneous UQ for triples of mediation effects (i.e. total, direct and indirect)}
        \2 \st{Can also do all 9 simultaneously}
    \1 UQ for predicted group-level effects? (not in this paper)
\end{outline}

@@

\begin{outline}
    \1 Monte Carlo study
        \2 Proof of concept
        \2 Explore robustness. See \citet{Sam23} for inspiration.
    \1 Real data
        \2 Trust study dataset? (this will be the focus of an applied paper)
\end{outline}
